---
title: "Data Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

Running a transformation audit requires three inputs. This vignette walks through
retrieving and cleaning each one so that joins work and the audit pipeline runs without
modification.

The three inputs are:

1. **Baseline data** — your analytic variable at the ZCTA level, retrieved from the
   Census ACS via `tidycensus`
2. **ZIP–ZCTA association table** — linking ZCTAs to their constituent ZIP codes,
   from the UDS Mapper crosswalk
3. **HUD ZIP–County crosswalk** — allocation ratios for ZIP-to-county redistribution,
   from HUD User

A note on column headers: the Census API, UDS Mapper, and HUD all deliver files with
different column naming conventions, inconsistent capitalization, and unsuffixed numeric
IDs that will silently fail on a join. The cleaning steps below are not optional. Every
join in the audit pipeline matches on specific standardized column names. If your columns
do not match, the join returns an empty table and the audit produces no output — with no
error to tell you why.

All code blocks are `eval = FALSE`. Replace the placeholders in capital letters with
your own values.

---

## 1. Census API Key

A free Census API key is required for ACS data retrieval via `tidycensus`. This is a
one-time setup per machine.

```{r api-key}
# Register at: https://api.census.gov/data/key_signup.html
# Then run this once — it writes the key to your .Renviron
tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)

# Restart R after running the above, then verify:
Sys.getenv("CENSUS_API_KEY")
```

---

## 2. Identify ZCTAs for Your County

The `zctaCrosswalk` package retrieves the ZCTAs associated with a target county using
the Census relationship file. This is the relationship-based membership definition —
the same ZCTA set used in ACS tabulations, and the correct baseline for the audit.

```{r zctas}
library(zctaCrosswalk)

# --- swap for your target county FIPS code ---
zctas <- get_zctas_by_county("27053")  # Hennepin County, MN
```

This is Decision Point 1 in the audit framework. Using geometric intersection instead of the
relationship file will change your baseline ZCTA count before any crosswalk is applied.
For Hennepin County, that difference is 20 ZCTAs — a 27% shift in the baseline prior
to any transformation. The audit measures perturbation from a fixed baseline; the
baseline definition is therefore a methodological choice, not a default.

---

## 3. Get Baseline ACS Data

```{r baseline}
library(shellgame)
library(tidycensus)
library(dplyr)
library(stringr)
library(janitor)

# --- swap variable code and year for your own ---
baseline_raw <- tidycensus::get_acs(
  geography = "zcta",
  variables = "B01001_001",  # total population; swap for any ACS variable
  year      = 2022,
  survey    = "acs5"
)

# Other commonly used variables:
# B19013_001 - Median household income
# B25001_001 - Total housing units
# B08201_001 - Households by vehicles available
```

### Clean the ACS output

The Census API returns a `GEOID` column containing the full ZCTA identifier with a
`"ZCTA5 "` prefix (e.g., `"ZCTA5 55401"`). The audit pipeline expects a plain 5-digit
`zcta` column. It also returns `estimate` and `moe` columns — keep only what you need.

```{r clean-baseline}
baseline_data <- baseline_raw |>
  janitor::clean_names() |>
  dplyr::filter(geoid %in% zctas) |>           # filter to your county's ZCTAs
  dplyr::mutate(
    zcta = stringr::str_pad(
      stringr::str_remove(geoid, "^ZCTA5 "),   # strip the prefix
      width = 5, pad = "0"                      # ensure zero-padded to 5 digits
    )
  ) |>
  dplyr::select(zcta, estimate)                 # keep only what the audit needs
```

**What can go wrong here:**

- If `geoid` still has the `"ZCTA5 "` prefix when you join, every match fails silently
- If `zcta` is stored as numeric rather than character, leading zeros are dropped and
  joins on ZCTAs beginning with `0` fail
- If you skip the `filter()`, your baseline will include ZCTAs outside your target
  county and inflate the baseline total

---

## 4. Prepare the ZIP–ZCTA Association Table

Download the UDS Mapper ZIP-to-ZCTA crosswalk from:
<https://www.udsmapper.org/zip-code-to-zcta-crosswalk/>

The file downloads as an Excel workbook. Export the relevant sheet to CSV before reading
it into R, or use `readxl::read_excel()` directly.

```{r zip-zcta}
library(readr)

# --- swap file path for your downloaded crosswalk ---
zip_zcta_raw <- readr::read_csv("path/to/ZiptoZCTA.csv", show_col_types = FALSE)
```

### Clean the ZIP–ZCTA crosswalk

The UDS Mapper file delivers columns named `ZIP_CODE` and `ZCTA` in all caps. The audit
pipeline expects lowercase `zip` and `zcta`. Both must be character class and
zero-padded to 5 digits.

```{r clean-zip-zcta}
zip_zcta <- zip_zcta_raw |>
  janitor::clean_names() |>                     # lowercases all column names
  dplyr::rename(
    zip  = zip_code,                            # check names(zip_zcta_raw) first
    zcta = zcta
  ) |>
  dplyr::mutate(
    zip  = stringr::str_pad(as.character(zip),  5, pad = "0"),
    zcta = stringr::str_pad(as.character(zcta), 5, pad = "0")
  ) |>
  dplyr::distinct() |>
  dplyr::filter(!is.na(zip), !is.na(zcta))
```

**What can go wrong here:**

- `ZIP_CODE` imported as numeric drops leading zeros on ZIPs beginning with `0`
  (New England, parts of the Mid-Atlantic) — joins will fail for those ZIPs
- Column names vary slightly across UDS Mapper vintages; run `names(zip_zcta_raw)`
  before assuming `zip_code` is the right name
- Duplicate rows are common and will inflate join results if not removed with
  `distinct()`

**Expected output structure:**

| Column | Type | Example |
|--------|------|---------|
| `zip`  | character | `"55401"` |
| `zcta` | character | `"55401"` |

---

## 5. Prepare the HUD ZIP–County Crosswalk

The HUD USPS ZIP–County crosswalk is derived from quarterly USPS vacancy data, which
records the distribution of business and residential mailing addresses across ZIP codes.
It is widely used in applied population and housing research because it provides
national coverage and allocates directly at the ZIP level. The `TOT_RATIO` field
represents the proportion of total addresses within a ZIP code associated with each
county.

HUD crosswalk files are directional and not invertible. The ZIP→County file allocates
from ZIP to county. It cannot be reversed — a separate County→ZIP file exists for the
opposite direction and the two cannot be used interchangeably. Select **ZIP to County**
when downloading.

Download from: <https://www.huduser.gov/portal/datasets/usps_crosswalk.html>

Select **ZIP to County** and the most recent available quarter.

```{r hud}
# --- swap file path for your downloaded crosswalk ---
hud_raw <- readr::read_csv("path/to/ZIP_COUNTY.csv", show_col_types = FALSE)
```

### Clean the HUD crosswalk

HUD delivers columns in all caps: `ZIP`, `COUNTY`, `TOT_RATIO`, `RES_RATIO`, etc.
The `COUNTY` column is a 5-digit FIPS code but is frequently imported as numeric,
dropping the leading zero for counties in states with FIPS codes beginning with `0`.

```{r clean-hud}
hud <- hud_raw |>
  janitor::clean_names() |>                     # lowercases all column names
  dplyr::mutate(
    zip    = stringr::str_pad(as.character(zip),    5, pad = "0"),
    county = stringr::str_pad(as.character(county), 5, pad = "0")
  ) |>
  dplyr::filter(!is.na(tot_ratio), tot_ratio > 0) |>
  dplyr::distinct()
```

### Choosing an allocation ratio — this is Decision Point 2

| Ratio | Description | Assumes |
|-------|-------------|---------|
| `tot_ratio` | Total addresses (default) | Total address count proxies variable distribution |
| `res_ratio` | Residential addresses only | Residential addresses are more appropriate proxy |
| `bus_ratio` | Business addresses | Business addresses are more appropriate proxy |
| `oth_ratio` | Other addresses | Other address types are more appropriate proxy |

Decision Point 2: Crosswalk Selection. Analysts must select among allocation ratios. Each
allocation ratio enacts a different set of assumptions on the underlying population.
The analyst who selects `tot_ratio` assumes that total addresses are a proxy for
population distribution; the analyst who selects `res_ratio` assumes residential
addresses are more appropriate. These are not technical details — they are
methodological choices that affect results.

To use a ratio other than `tot_ratio`, retain the desired column and rename it
accordingly before passing to `geoDeltaAudit`.

**What can go wrong here:**

- `county` imported as numeric drops the leading zero for states with FIPS `< 10000`
  (Alabama, Alaska, Arizona, Arkansas, California, Colorado, Connecticut) — every
  county-level join fails for those states
- Rows with `tot_ratio == 0` exist in the HUD file and should be filtered before use;
  they represent ZIP codes with no address activity in that county
- HUD column names are consistent across quarters but the filename changes with each
  release — check `names(hud_raw)` if `clean_names()` produces unexpected results

**Expected output structure:**

| Column | Type | Example |
|--------|------|---------|
| `zip` | character | `"55401"` |
| `county` | character | `"27053"` |
| `tot_ratio` | numeric | `0.843` |

---

## 6. Verify Before Handing Off

Run these checks before passing data to `geoDeltaAudit`. They take thirty seconds and
will prevent hours of debugging a silent empty-join result.

```{r verify}
# All ID columns should be character class, 5 digits, no NAs
stopifnot(is.character(baseline_data$zcta))
stopifnot(all(nchar(baseline_data$zcta) == 5))
stopifnot(!any(is.na(baseline_data$zcta)))

stopifnot(is.character(zip_zcta$zip))
stopifnot(is.character(zip_zcta$zcta))

stopifnot(is.character(hud$zip))
stopifnot(is.character(hud$county))

# ZCTAs in baseline should match ZCTAs in association table
matched <- sum(baseline_data$zcta %in% zip_zcta$zcta)
cat("ZCTAs matched to association table:", matched, "of", nrow(baseline_data), "\n")
```

If `matched` is zero, the join will produce an empty result. Go back and check padding
and class on both sides before proceeding.

---

## Next Steps

Once all three inputs are cleaned and verified, pass them to `geoDeltaAudit` to run
the audit:

- `vignette("audit_walkthrough", package = "geoDeltaAudit")` — step-by-step execution
  with copy-paste swappable chunks
- `vignette("conceptual_framework", package = "shellgame")` — what the audit measures
  and why crosswalk decisions matter
